---
title: "ITS_soil"
author: "Emily Giroux"
date: "10/31/2021"
output: pdf_document
fontsize: 11pt
geometry: margin=1in
urlcolor: blue
header-includes: \usepackage{xcolor}
---

```{r setup, include=FALSE}
#Set the global options for knitr
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy = TRUE, fig.align='center',
               cache=FALSE, collapse=TRUE, echo=FALSE, eval=FALSE, include=FALSE,
               message=FALSE, quietly=TRUE, results='hide', warn.conflicts=FALSE, 
               warning=FALSE)
```

**Using package `BiocManager` to install required packages:**
```{r, biocInstall, eval=TRUE, echo=FALSE, include=TRUE, cache=TRUE, tidy=FALSE, message=FALSE}
r <- getOption("repos")
r["CRAN"] <- "http://cran.us.r-project.org"
options(repos = r)
if (!requireNamespace("BiocManager"))
    install.packages("BiocManager")
BiocManager::install()
library("BiocManager")
.cran_packages <- c("cowplot", "data.table", "ggplot2", "knitr", "rprojroot")
.bioc_packages <- c("BiocStyle", "Biostrings", "dada2", "microbiome", "phyloseq",
                    "ShortRead")
.inst <- .cran_packages %in% installed.packages()
if(any(!.inst)) {
   install.packages(.cran_packages[!.inst])
}
.inst <- .bioc_packages %in% installed.packages()
if(any(!.inst)) {
  BiocManager::install(.bioc_packages[!.inst], ask = FALSE)
}
```

**Load packages into session, and print package versions:**
```{r, showBiocPackages, echo=FALSE, eval=TRUE, include=TRUE, results='hold', cache=TRUE}
sapply(c(.cran_packages, .bioc_packages), require, character.only = TRUE)
```

Source our custom R scripts:    
For this we will use the rprojroot package to set the directory structures. This will help us when finding our files to source functions. We specify ours is an RStudio project. The root object contains a function that will help us locate our package R files regarless of our current working directory.
```{r sourcing_my_functions, echo=FALSE, eval=TRUE, include=FALSE, cache=TRUE}
library("rprojroot")
root <- rprojroot::is_rstudio_project
scriptsPath <- root$make_fix_file(".")("R")
scripts  <- dir(root$find_file("R", path = root$find_file()))
scriptsl <- paste(scriptsPath, scripts, sep = "//")
lapply(scriptsl, source)
```

Record the path to the environment images directory:
```{r importChptImage, echo=TRUE, eval=TRUE, include=FALSE, cache=TRUE}
sharedPath <- "/isilon/cfia-ottawa-fallowfield/users/girouxeml/PIRL_working_directory/"
analysis <- "ecobiomics/"
sharedPathAn <- paste(sharedPath, analysis, sep = "")
imageDirPath <- "/home/CFIA-ACIA/girouxeml/GitHub_Repos/r_environments/ecobiomics/"
```

Recall from Chapter 2b ITS DADA2-specific processing step: The only variables we need from the previous environment for this analysis chapter are the "ps" variable from the fitGTR image, which we re-saved in the chptImageB variable:
```{r}
chptImageB <- "ecobiomics_ITS_2b.RData"
load(paste(imageDirPath, chptImageB, sep = ""))
```

Fresh image for this new chapter, and image to continue saving going forward:
```{r}
chptImage    <- "ecobiomics_ITS_analysis_soil.RData"
save.image(paste(imageDirPath, chptImage, sep = ""))
```

# Importing required data:    
# 1. ps variable from ITS DADA2 processing chapter "loadingChptImage"        
# 2. Import sample metadata "importingSampleMetadata"     


Pre-processing data for analysis:    
1. Extract ITS data specific to Soil samples "subsetSoilPSdata"    
2. Assign ASV names to ASV DNA sequences for readability "assignASVshortNames"    
    
    
Fungi in Soil Samples Across Extraction Kits
```{r subsetSoilPSdata}
psSoil <- subset_samples(ps, Sample=="soil")
```
It is more convenient to use short names for our ASVs (e.g. ASV21) rather than the full DNA sequence when working with some of the tables and visualizations from phyloseq, but we want to keep the full DNA sequences for other purposes like merging with other datasets or indexing into reference databases like the Earth Microbiome Project. For that reason we’ll store the DNA sequences of our ASVs in the refseq slot of the phyloseq object, and then rename our taxa to a short string. That way, the short new taxa names will appear in tables and plots, and we can still recover the DNA sequences corresponding to each ASV as needed with refseq(ps).
```{r assignASVshortNames}
dna <- Biostrings::DNAStringSet(taxa_names(psSoil))
names(dna) <- taxa_names(psSoil)
psSoil <- merge_phyloseq(psSoil, dna)
taxa_names(psSoil) <- paste0("ASV", seq(ntaxa(psSoil)))
psSoil
```
```{r}
library("ggplot2")
dir.create(paste(sharedPathReg, "ggplots", sep = ""), showWarnings = TRUE, recursive = FALSE)
ggplotsPath <- paste(sharedPathReg, "ggplots", sep = "")

png(file = paste(ggplotsPath, "/alphaDiversity_ITS_Soil", ".png", sep = ""))
alphaDiversity <- plot_richness(psSoil, x="ExtractionKit", measures=c("Shannon", "Simpson"), color="Sample",
                                title = "Alpha diversity of ITS soil samples by extraction kit")
alphaDiversity + theme(plot.title = element_text(hjust = 0.5))
print(alphaDiversity + theme(plot.title = element_text(hjust = 0.5)))
dev.off()

# Visualize alpha-diversity:
alphaDiversity <- plot_richness(psSoil, x="ExtractionKit", measures=c("Shannon", "Simpson"), color="Sample",
                                title = "Alpha diversity of ITS soil samples by extraction kit")
alphaDiversity + theme(plot.title = element_text(hjust = 0.5))
```

Here is if we filter based on having to know the species:
```{r, testCmds2, eval=FALSE, include=FALSE, echo=-5, message=FALSE, comment=NA}
ps2Species <- subset_taxa(psSoil, !is.na(Species) & !Species %in% c("", "uncharacterized"))
rank_names(ps2Species)
table(tax_table(ps2Species)[, "Species"], exclude = NULL)
```

Generate a table that shows the prevalence of species present across Soil samples:    
Prevalence in the dataset is defined here as the number of samples in which a taxa appears at least once.
```{r}
prevDfSoil <- apply(X = otu_table(ps2Species),
                     MARGIN = ifelse(taxa_are_rows(ps2Species), yes = 1, no = 2),
                     FUN = function(x){sum(x>0)})
prevDfSoil <- data.frame(Prevalence = prevDfSoil,
                          TotalAbundance = taxa_sums(ps2Species),
                          tax_table(ps2Species))
# Are there species that are comprised of mostly low-prevalence features? Compute the total and average prevalences of the features in each species.
prevTblSppSoil <- plyr::ddply(prevDfSoil, "Species", function(df1){cbind(mean(df1$Prevalence), sum(df1$Prevalence), mean(df1$Prevalence)/nsamples(ps2Species)*100)})
colnames(prevTblSppSoil) <- c("Species", "Mean", "TotalASVprevalence", "PercentOfSamples")
prevTblSppSoil
prevTblSppSoil <- plyr::ddply(prevDfSoil, "Species", function(df1){cbind(mean(df1$Prevalence), 
                                                                         sum(df1$Prevalence), 
                                                                         sum(df1$Prevalence)/mean(df1$Prevalence), 
                                                                         mean(df1$Prevalence)/sum(df1$Prevalence)*nsamples(ps2Species))})
colnames(prevTblSppSoil) <- c("Species", "MeanASVAcrossSamples", "TotalPrevalence", "TotalASVInData", "NumberSamplesWithASV")
head(prevTblSppSoil)
prevTblSppSoil[order(-prevTblSppSoil$NumberSamplesWithASV),]
```

Remove low-abundance species that appear in less than 10% of the mean number of samples by sample type: 
```{r}
library("dplyr")
# How many samples represent 10% of total number of samples?
nsamples(ps2Species) # 15
unique(sample_data(ps2Species)$ExtractionKit)
x <- aggregate(data.frame(SampleSize = sample_data(ps2Species)$ExtractionKit), list(ExtractionKit = sample_data(ps2Species)$ExtractionKit), length)
prevFilter <- mean(x$SampleSize)*0.1

# When getting prevalence filter, set it to be the number that is 10% of the mean number of samples by sample type
arrange(filter(prevTblSppSoil, TotalPrevalence > prevFilter))

# Filter-out low-abundance species that appear in less than the value of the prevalence fitler "prevFilter" calculated above:
spps <- subset(prevTblSppSoil, prevTblSppSoil$TotalPrevalence > prevFilter)
ps2SpeciesFiltered  <- subset_taxa(ps2Species, Species %in% spps$Species)
rank_names(ps2SpeciesFiltered)
table(tax_table(ps2SpeciesFiltered)[, "Species"])
tax_table(ps2SpeciesFiltered)
```


Taxa prevalence versus total counts after filtering. Each point is a different taxa. Exploration of the data in this way is often useful for selecting filtering parameters, like the minimum prevalence criteria we will used to filter the data above.
```{r}
library(ggplot2)
prevdf1 = subset(prevDfSoil, Species %in% get_taxa_unique(ps2SpeciesFiltered, "Species"))
ggplot(prevdf1, aes(TotalAbundance, Prevalence / nsamples(psSoil),color=Species)) +
 # Include a guess for parameter
 geom_hline(yintercept = 0.05, alpha = 0.5, linetype = 2) + geom_point(size = 2, alpha = 0.7) +
 scale_x_log10() + xlab("Total Abundance") + ylab("Prevalence [Frac. Samples]") +
 facet_wrap(~Species) + theme(legend.position="none")
```

```{r}
head(taxa_names(ps2SpeciesFiltered))
head(tax_table(ps2SpeciesFiltered)[,7])
table(tax_table(ps2SpeciesFiltered)[,7])
```

How many species would be present after filtering?
```{r}
length(get_taxa_unique(ps2SpeciesFiltered, taxonomic.rank = "Species"))
```
Now lets filter out samples (outliers and low performing samples)
Do some simple ordination looking for outlier samples, first we variance stabilize the data with a log transform, then perform PCoA using bray’s distances
```{r}
logt          <- phyloseq::transform_sample_counts(ps2SpeciesFiltered, function(x) log(1 + x) )
out.pcoa.logt <- phyloseq::ordinate(logt, method = "PCoA", distance = "bray")
evals         <- out.pcoa.logt$values$Eigenvalues
phyloseq::plot_ordination(logt, out.pcoa.logt, type = "Sample", 
                          color = "ExtractionKit") + ggplot2::labs(col = "ExtractionKit") +
  coord_fixed(sqrt(evals[2] / evals[1]))
```

```{r}
# Visualize alpha-diversity:
library("ggplot2")

png(file = paste(ggplotsPath, "/alphaDiversity_corrected_ITS_Soil", ".png", sep = ""))
alphaDiversityCorrected <- plot_richness(logt, x="ExtractionKit", measures=c("Shannon", "Simpson"), color="Sample",
                                         title = "Alpha diversity of ITS soil samples by extraction kit")
alphaDiversityCorrected + theme(plot.title = element_text(hjust = 0.5))
print(alphaDiversityCorrected + theme(plot.title = element_text(hjust = 0.5)))
dev.off()

# Visualize alpha-diversity:
alphaDiversityCorrected <- plot_richness(logt, x="ExtractionKit", measures=c("Shannon", "Simpson"), color="Sample",
                                         title = "Alpha diversity of ITS soil samples by extraction kit")
alphaDiversityCorrected + theme(plot.title = element_text(hjust = 0.5))
```
```{r}
# Transform data to proportions as appropriate for Bray-Curtis distances
ps.prop <- transform_sample_counts(logt, function(otu) otu/sum(otu))
ord.nmds.bray <- ordinate(ps.prop, method="NMDS", distance="bray")
```
```{r}
plot_ordination(ps.prop, ord.nmds.bray, color="ExtractionKit", title="Bray NMDS")
```




```{r}
# Bar plot:
top20 <- names(sort(taxa_sums(logt), decreasing=TRUE))[1:36] # Some unique asvs may group together at the species level and thus decrease the top 20 count, so manually increase this until 20 species get included in the abundance graph.
ps.top20 <- transform_sample_counts(logt, function(OTU) OTU/sum(OTU))
ps.top20 <- prune_taxa(top20, ps.top20)
```

```{r}
library("phyloseq")
library("ggplot2")
png(file = paste(ggplotsPath, "/top20SpeciesAbundance_ITS_Soil", ".png", sep = ""))
sppAbundanceTop20 <- plot_bar(ps.top20, x = "ExtractionKit", fill = "Species",
                              title = "Abundance of top 20 species for ITS soil samples by extraction kit",
                              facet_grid = ~sample_Sample) +
  geom_bar(aes(color = Species, fill = Species), stat = "identity", position = "stack") +
  guides(fill=guide_legend(ncol=2)) # This splits the legend into 2 columns for readability.
print(sppAbundanceTop20)
dev.off()

sppAbundanceTop20
```


```{r}
library("phyloseq")
library("ggplot2")

df <- psmelt(logt)
df$group <- paste0(df$Phylum, "-", df$Family, sep = "")# Need to fix the group names by removing the extra characters in the names
df$group <- sub("p__", "", df$group)
df$group <- sub("f__", "", df$group)
df$ScientificName <- paste0(df$Genus, "-", df$Species, sep = "")
df$ScientificName <- sub("g__", "", df$ScientificName)
df$ScientificName <- sub("s__", "", df$ScientificName)
df$ScientificName <- sub("-", "_", df$ScientificName)
df$Kingdom <- sub("k__", "", df$Kingdom)
df$Phylum <- sub("p__", "", df$Phylum)
df$Class <- sub("c__", "", df$Class)
df$Order <- sub("o__", "", df$Order)
df$Family <- sub("f__", "", df$Family)
df$Genus <- sub("g__", "", df$Genus)
df$Species <- sub("s__", "", df$Species)

df %>% dplyr::count(Phylum) %>% arrange(desc(n))
# Therefore, in the final phyloseq, the most abundant phyla fall under Ascomycota, Basidiomycota, and Mortierellomycota, and Chytridiomycota, others are Mucoromycota, Glomeromycota, Kickxellomycota, Blastocladiomycota, Basidiobolomycota, Monoblepharomycota.
```


```{r}
df2 <- psmelt(ps.top20)
df2 %>% dplyr::count(Phylum) %>% arrange(desc(n))
# Therefore, in the top 20 species, the phyla fall under Ascomycota, Basidiomycota, and Mortierellomycota.

df2$group <- paste0(df2$Phylum, "-", df2$Family, sep = "")
df2$group <- sub("p__", "", df2$group)
df2$group <- sub("f__", "", df2$group)
df2$ScientificName <- paste0(df2$Genus, "-", df2$Species, sep = "")
df2$ScientificName <- sub("g__", "", df2$ScientificName)
df2$ScientificName <- sub("s__", "", df2$ScientificName)
df2$ScientificName <- sub("-", "_", df2$ScientificName)
```

Abundance of top 20 species plot, using df2:
```{r}
png(file = paste(ggplotsPath, "/top20SpeciesAbundanceSciNames_ITS_Soil", ".png", sep = ""))
sppAbundanceTop20_SciNames <- ggplot(df2, aes_string(x = "ExtractionKit", y = "Abundance", fill = "ScientificName"),
                                     title = "Abundance of top 20 species for ITS soil samples by extraction kit",
                                     facet_grid = ~sample_Sample) +
  geom_bar(aes(color = ScientificName, fill = ScientificName), stat = "identity", position = "stack") +
  ggtitle("Abundance of top 20 species for ITS \n soil samples by extraction kit") + 
  theme(plot.title = element_text(hjust = 0.5)) +
  guides(col=guide_legend(ncol=2)) # This splits the legend into 2 columns for readability.
print(sppAbundanceTop20_SciNames)
dev.off()
```

```{r}
sppAbundanceTop20_SciNames
```
```{r}
phylums <- c("p__Ascomycota", "p__Basidiomycota", "p__Chytridiomycota", "p__Mortierellomycota")

df3 <- df
df3$Phylum[!df3$Phylum %in% phylums] <- "Others"
df3$Family[!df3$Phylum %in% phylums] <- "Others"

df3$Family[df3$Phylum=="p__Ascomycota" & 
 !df3$Family %in% c("f__Helotiaceae", "f__Hypocreaceae", "f__Myrmecridiaceae", "f__Lasiosphaeriaceae", "f__Lindgomycetaceae", "f__Nectriaceae")] <- "Other Ascomycota"

df3$Family[df3$Phylum=="p__Basidiomycota" &
 !df3$Family %in% c("f__Erythrobasidiaceae", "f__Mrakiaceae", "f__Malasseziaceae", "f__Ustilaginaceae")] <- "Other Basidiomycota"

df3$Family[df3$Phylum=="p__Mortierellomycota" & 
 !df3$Family %in% c("f__Mortierellaceae")] <- "Other Mortierellomycota"

df3$Family[df3$Phylum=="p__Chytridiomycota" & 
 !df3$Family %in% c("f__Rhizophydiales_fam_Incertae_sedis", "f__Rhizophlyctidaceae", "f__Spizellomycetaceae")] <- "Other Chytridiomycota"
```

Using df with edited Kingdom, Phylum, Class, etc edited:
```{r}
phylums <- c("Ascomycota", "Basidiomycota", "Chytridiomycota", "Mortierellomycota")

df3 <- df
df3$Phylum[!df3$Phylum %in% phylums] <- "Others"
df3$Family[!df3$Phylum %in% phylums] <- "Others"

df3$Family[df3$Phylum=="Ascomycota" & 
 !df3$Family %in% c("Helotiaceae", "Hypocreaceae", "Myrmecridiaceae", "Lasiosphaeriaceae", "Lindgomycetaceae", "Nectriaceae")] <- "Other Ascomycota"

df3$Family[df3$Phylum=="Basidiomycota" &
 !df3$Family %in% c("Erythrobasidiaceae", "Mrakiaceae", "Malasseziaceae", "Ustilaginaceae")] <- "Other Basidiomycota"

df3$Family[df3$Phylum=="Mortierellomycota" & 
 !df3$Family %in% c("Mortierellaceae")] <- "Other Mortierellomycota"

df3$Family[df3$Phylum=="Chytridiomycota" & 
 !df3$Family %in% c("Rhizophydiales_fam_Incertae_sedis", "Rhizophlyctidaceae", "Spizellomycetaceae")] <- "Other Chytridiomycota"

library("forcats")
library("dplyr")
df4 <- select(df3, ExtractionKit, Phylum, Family) %>%
  mutate(Phylum=factor(Phylum, levels=c(phylums, "Others")),
         Family=fct_reorder(Family, 10*as.integer(Phylum) + grepl("Others", Family))) %>%
#  group_by(Family)
  group_by(Family) %>%  # For this dataset only
  sample_n(3)         # Otherwise, unnecessary
```

```{r}
ColourPalleteMulti <- function(df, group, subgroup){

  # Find how many colour categories to create and the number of colours in each
  categories <- aggregate(as.formula(paste(subgroup, group, sep="~" )), df, function(x) length(unique(x)))
  category.start <- (scales::hue_pal(l = 100)(nrow(categories))) # Set the top of the colour pallete
  category.end  <- (scales::hue_pal(l = 40)(nrow(categories))) # set the bottom

  # Build Colour pallette
  colours <- unlist(lapply(1:nrow(categories),
                          function(i){
                            colorRampPalette(colors = c(category.start[i], category.end[i]))(categories[i,2])}))
  return(colours)
}
```

```{r}
colours <- ColourPalleteMulti(df4, "Phylum", "Family")
```

```{r}
library(ggplot2)
ggplot(df4, aes(x=ExtractionKit, fill = Family)) + 
  geom_bar(position="fill", colour = "grey") +  # Stacked 100% barplot
  scale_fill_manual("", values=colours) +
  theme(axis.text.x=element_text(angle=90, vjust=0.5)) +  # Vertical x-axis tick labels
  scale_y_continuous(labels = scales::percent_format()) +
  labs(y="Relative abundance")
```
```{r}
coord_fam = df4 %>% select(Phylum, Family) %>% unique()  %>% ungroup()%>%mutate(x= c(rep(1,nrow(.))), y=1:nrow(.))

coord_phylum = coord_fam %>% group_by(Phylum) %>% summarise(x=mean(x),ymid= mean(y),
                                                           ymin=min(y), ymax=max(y))

v=0.3
p2 = coord_fam %>% ggplot()+
  geom_point(aes(0.05,y, col= Family), size=8 )+
  scale_x_continuous(limits = c(0, 2)) +
  geom_segment(data = coord_phylum,
               aes(x = x + 0.1, xend = x + v, y= ymax, yend=ymax), col="black")+
  
  geom_segment(data = coord_phylum,
               aes(x = x + 0.1, xend = x + v, y= ymin, yend=ymin))+
  
  geom_segment(data = coord_phylum,
               aes(x = x + v, xend = x + v, y= ymin, yend=ymax))+
  
  geom_text(data = coord_phylum, aes(x = x + v+0.5, y = ymid, label = Phylum)) +
  geom_text(data = coord_fam, aes( x=0.6, y=y, label=Family, col=Family))+
  geom_text(data = coord_fam, aes( x=0.6, y=y, label=Family), alpha=0.9,col="grey50")+
  scale_colour_manual(values = colours)+
  theme_void()+theme(legend.position = "none")+ 
  scale_y_reverse()

p2
```

```{r}
#install.packages("patchwork")
library("patchwork")

p1 <- ggplot(df4, aes(x=ExtractionKit, fill = Family)) + 
  geom_bar(position="fill", colour = "grey", show.legend = FALSE) +  # Stacked 100% barplot
  scale_fill_manual("", values=colours) +
  theme(axis.text.x=element_text(angle=90, vjust=0.5)) +  # Vertical x-axis tick labels
  scale_y_continuous(labels = scales::percent_format()) +
  labs(y="Relative abundance")


p1+p2 + plot_layout(ncol=2, heights = c(100,1), widths = c(1,2))
png(file = paste(ggplotsPath, "/AbundanceMainPhylumsFamilies_ITS_Soil", ".png", sep = ""))
print(p1+p2 + plot_layout(ncol=2, heights = c(100,1), widths = c(1,2)))
dev.off()
```



Out of curiosity how many "reads" does this leave us at???
```{r}
sum(colSums(otu_table(psSoil))); sum(colSums(otu_table(ps2Species))); sum(colSums(otu_table(ps2SpeciesFiltered))); sum(colSums(otu_table(logt)))
```
```{r}
library("microbiome")
library("phyloseq")
summarize_phyloseq(logt)
table(sample_data(psSoil)$ExtractionKit)
#FastDNA  PowerSoil PowerWater 
#         5          6          4 
table(sample_data(logt)$ExtractionKit)
#FastDNA  PowerSoil PowerWater 
#         5          6          4

sample_data(logt)$totalReads <- sample_sums(logt)
meta(logt)[,c("ExtractionKit", "totalReads")]
readsSamplesTbl <- meta(logt)[,c("ExtractionKit", "totalReads")]
fastDNAReads <- readsSamplesTbl[which(readsSamplesTbl$ExtractionKit=="FastDNA"),]
powerSoilReads <- readsSamplesTbl[which(readsSamplesTbl$ExtractionKit=="PowerSoil"),]
powerWtrReads <- readsSamplesTbl[which(readsSamplesTbl$ExtractionKit=="PowerWater"),]
sum(fastDNAReads$totalReads) # [1] 6909.629
sum(powerSoilReads$totalReads) # [1] 2984.067
sum(powerWtrReads$totalReads) # [1] 1772.843

extrkits <- unique(sample_data(logt)$ExtractionKit)
extrkitlogtTbl <- table(sample_data(logt)$ExtractionKit)
rbind(extrkits, extrkitlogtTbl)
extrkitlogtTbl$ExtractionKit <- 

library(tidyverse)
library(dplyr)
sample_data(logt) %>% group_by(ExtractionKit) %>% dplyr::summarise((n = length(unique(totalReads))))

```


Below I'm testing what a heatmap would look like for taxa abundance across extraction kits:
```{r, plotHeatMap, echo=TRUE, eval=TRUE, cache=TRUE, tidy=FALSE, message=FALSE, comment=NA, warning=FALSE}
library(phyloseq)
library(ggplot2)
plot_heatmap(ps.top20, "PCoA", distance="bray", 
             sample.label="ExtractionKit", 
             taxa.label="Species",
             low="#66CCFF", high="#000033", na.value="white") +
  scale_x_discrete(expand=c(0,0))
```








IF we go by Genus level instead of Species:
Here is if we filter based on having to know the genus:
```{r, testCmds2, eval=FALSE, include=FALSE, echo=-5, message=FALSE, comment=NA}
ps2Genus <- subset_taxa(psSoil, !is.na(Genus) & !Genus %in% c("", "uncharacterized"))
rank_names(ps2Genus)
table(tax_table(ps2Genus)[, "Genus"], exclude = NULL)
```

Generate a table that shows the prevalence of genera present across Soil samples:    
Prevalence in the dataset is defined here as the number of samples in which a taxa appears at least once.
```{r}
prevDfSoil <- apply(X = otu_table(ps2Genus),
                     MARGIN = ifelse(taxa_are_rows(ps2Genus), yes = 1, no = 2),
                     FUN = function(x){sum(x>0)})
prevDfSoil <- data.frame(Prevalence = prevDfSoil,
                          TotalAbundance = taxa_sums(ps2Genus),
                          tax_table(ps2Genus))
# Are there genera that are comprised of mostly low-prevalence features? Compute the total and average prevalences of the features in each genus.
prevTblGenSoil <- plyr::ddply(prevDfSoil, "Genus", function(df1){cbind(mean(df1$Prevalence), sum(df1$Prevalence), mean(df1$Prevalence)/nsamples(ps2Genus)*100)})
colnames(prevTblGenSoil) <- c("Genus", "Mean", "TotalASVprevalence", "PercentOfSamples")
prevTblGenSoil
prevTblGenSoil <- plyr::ddply(prevDfSoil, "Genus", function(df1){cbind(mean(df1$Prevalence), 
                                                                         sum(df1$Prevalence), 
                                                                         sum(df1$Prevalence)/mean(df1$Prevalence), 
                                                                         mean(df1$Prevalence)/sum(df1$Prevalence)*nsamples(ps2Genus))})
colnames(prevTblGenSoil) <- c("Genus", "MeanASVAcrossSamples", "TotalPrevalence", "TotalASVInData", "NumberSamplesWithASV")
head(prevTblGenSoil)
prevTblGenSoil[order(-prevTblGenSoil$NumberSamplesWithASV),]
```

Remove low-abundance genera that appear in less than 10% of the mean number of samples by sample type: 
```{r}
library("dplyr")
# How many samples represent 10% of total number of samples?
nsamples(ps2Genus) # 39
unique(sample_data(ps2Genus)$ExtractionKit)
x <- aggregate(data.frame(SampleSize = sample_data(ps2Genus)$ExtractionKit), list(ExtractionKit = sample_data(ps2Genus)$ExtractionKit), length)
prevFilter <- mean(x$SampleSize)*0.1
# When getting prevalence filter, set it to be the number that is 10% of the mean number of samples by sample type
arrange(filter(prevTblGenSoil, TotalPrevalence > prevFilter))
# Filter-out low-abundance genera that appear in less than the value of the prevalence fitler "prevFilter" calculated above:
genera <- subset(prevTblGenSoil, prevTblGenSoil$TotalPrevalence > prevFilter)
ps2GenusFiltered    <- subset_taxa(ps2Genus, Genus %in% genera$Genus)
rank_names(ps2GenusFiltered)
table(tax_table(ps2GenusFiltered)[, "Genus"])
tax_table(ps2GenusFiltered)
```

Taxa prevalence versus total counts after filtering. Each point is a different taxa. Exploration of the data in this way is often useful for selecting filtering parameters, like the minimum prevalence criteria we will used to filter the data above.
```{r}
prevdf1 = subset(prevDfSoil, Genus %in% get_taxa_unique(ps2GenusFiltered, "Genus"))
ggplot(prevdf1, aes(TotalAbundance, Prevalence / nsamples(psSoil),color=Genus)) +
 # Include a guess for parameter
 geom_hline(yintercept = 0.05, alpha = 0.5, linetype = 2) + geom_point(size = 2, alpha = 0.7) +
 scale_x_log10() + xlab("Total Abundance") + ylab("Prevalence [Frac. Samples]") +
 facet_wrap(~Genus) + theme(legend.position="none")
```

```{r}
head(taxa_names(ps2GenusFiltered))
head(tax_table(ps2GenusFiltered)[,6])
table(tax_table(ps2GenusFiltered)[,6])
```

How many genera would be present after filtering?
```{r}
length(get_taxa_unique(ps2GenusFiltered, taxonomic.rank = "Genus"))
```